\documentclass{article}
\usepackage{gamesem}
\usepackage{amsmath}

\title{TODO list}
\author{William Blum}
\begin{document}
\maketitle

\section{Space complexity}
\begin{enumerate}

\item Express the size of $\eta$-long expansion with respect to the size of the original term. Probably something like: $|\eta\mbox{-nf}(M:\tau)| = \mathcal{O}(|M|.|\tau|)$, which would mean a size polynomial in the size of the input (=size of term + size of its type).

\item In fact beta-normalization is PSPACE(-complete) at order 3! This means that although by Sec.~\ref{sec:misc}:\ref{safe_varnames} there are order-3 terms needing arbitrarily large number of variables names, in order to reduce any given order 3 terms, we only need to introduce a number of fresh variable names that is at most exponential in the size of the formula (since encoding a variable name requires space logarithmic in the total number of names).

\item At order 3, it should be possible to prove the PSPACE complexity result by showing that the maximum size of the stack of the DPDA encoding the game semantics of the term (see Lukes' paper on observation equivalence of order-3 finitary IA) is polynomial in the size of the term.


\item With safety, no names need to be introduced at any order. Thus in a certain sense we save some space, but only logarithmic in the number of variable names that we would have needed to introduced without the safety constraint.

\item When the order is not bound, normalization of safe term is non-elementary.
But at this level ``TIME=SPACE''. So safety probably does not give anything in term of space complexity.

\item The last two points raise the following question: When the order is bound, does it really cost anything to introduce fresh variable names? For instance, is it possible to find say an order-$4$ (unsafe) term for which an unbounded number of variable names need to be introduced in order to be able to reduce it (using any possible reduction strategy)?
    No answer at the moment.

\item I expect the complexity of normalization for order-$r$ simply-typed terms where $r\geq4$ to be something like $(r-3)$-EXPSPACE.
    What would safety give use in that case?

\item Space complexity of beta normalization at order 4 for safe terms?
EXPSPACE? Safety allows us to not have to encode variable names.

\item Space complexity of beta eta equality at order 4 for safe terms?

\end{enumerate}

\section{Miscellaneous ideas}
\label{sec:misc}
\begin{enumerate}

\item A general idea to develop: ``safety appears at order 3 but kicks in at order 4''. For instance the simplest closed beta normal form is order 3 (kierstead term). But at order 3, the economy of variable names does not buy you anything since normalization is already PSPACE-complete! In fact we hope to prove that safety at order 4 causes a gain in term of space complexity for the normalization problem.
    Moreover I suspect that De Miranda's result on the nonnecessity of safety for generation word languages at order 2 can be extended to order 3.


\item find a generic encoding of any non-elementary time turing machine
in the safe lambda calculus (using a non-uniform representation of
numbers)


\item \label{safe_varnames} PB: How many variable names are required to generate all the
closed beta normal form of a given type?

In the lambda calculus, it is unbounded. Take the order-3 type
$$\tau
\cong (((o,o),o), (o^n\typear o), o, o) \ . $$

For any $n\in\nat$ the following beta normal term:
$$ M \cong \lambda f g x . f (\lambda x_1.f(\lambda x_2.f(\lambda \ldots f(\lambda \ldots f(\lambda x_{n-1}.g x_1 \ldots x_{n-1}) \ldots )$$
is of type $\tau$ and uses $n$ distinct variable names. It is easy to see that
any other $\beta$-normal form using less that $n$ distinct names cannot be alpha converted to $M$
since the function $g$ expect as argument $n$ variables bound at $n$ different places.


In the safe lambda calculus, however, the number of variable names
required to generate all the closed beta normal forms is bounded by
$size(\tau)-1$ where $size$ is defined by $size(o)= 1$ and
$size(\tau\typear\mu) = size(\tau) + size(\mu)$.


\item In a call-by-value programming language, safety does not buy you anything because the reduction strategy is such that no reduction occurs under a lambda abstraction. Consequently, when evaluating a (closed) program, $\beta$-reduction will never require variable renaming!

    There is still a potential, however, for applications in the domain of partial evaluation. Indeed, partial evaluation strategies proceed by contracting redex within the body of a lambda abstraction. The safety constraint allows you to perform these reductions without having to introduce new variable names.
\end{enumerate}

\section{Type inhabitation}

Let us say that a type is UNSAFE if it is inhabited, and every
inhabitant is unsafe; further we say that an unsafe type is UNIQUELY
UNSAFE if it has a unique inhabitant.

\begin{enumerate}

\item All types with one atom are safe (either not inhabited at all or inhabited in the safe lambda calculus). What about multiple atoms?

For any atom $a$, we can construct the types $\tau(n,a)$
as follows: $\tau(0,a) = a$ and $\tau(n+1,a) = \tau(n,a) \typear a$.
It is easy to see that for all $k\in\nat$ and atom $a$, $\tau(2k,a)$ is not inhabited by any term (even unsafe) and $\tau(2k+1,a)$ is safely inhabited.

Take three distinct atoms $a$, $b$ and $c$. For any $i,j,k\in\nat$, we write $\sigma(i,j,k)$ to denote the type
$$\sigma(i,j,k) \cong (\tau(i,a) \typear \tau(j,b)) \typear (\tau(j,b) \typear \tau(k,c)) \typear \tau(i,a) \typear
\tau(k,c) \ .$$
We now give examples of types together with some inhabitant (safe if one is known):
\begin{enumerate}
\item $\sigma(i,j,k)$ is inhabited by the term $\lambda x y z w . y (x z)$ which is unsafe if and only if $i<j$.

\item For any odd number $k$, $\sigma(i,j,k)$ is safely inhabited. Indeed if $k=1$ then it is inhabited by the safe term
    $$\lambda x y z w^c. w$$
    and otherwise it is inhabited by
    $$\lambda x y z w^{\tau(c,k-1)}. w (In(k-2))$$
    where $In(k-2)$ denotes the safe inhabitant of type $\tau(c,k-2)$ (which exists since $k-2$ is odd.

\item The single-atom type
$(\tau(1,a) \typear \tau(2,a)) \typear (\tau(2,a) \typear \tau(4,a)) \typear \tau(1,a) \typear
\tau(4,a)$ is inhabited by the safe term $$\lambda x^{\tau(1,a)
\typear \tau(2,a)} y^{(\tau(2,a) \typear \tau(4,a))} z^{\tau(1,a)}
w^{\tau(3,a)} . x (\lambda x^a . x) (\lambda x^a . x)\ .$$

\item $\sigma(1,2,4)$
is inhabited by the safe term:
$$ \lambda x y z w . y (\lambda g . x (\lambda a . a) (\lambda b . b)) (\lambda f . f (\lambda c . c))$$
where
$x : ((a, a), (b, b), b)$,
$y : (((b, b), b), (((c, c), c), c), c)$,
$g : (b, b)$,
$f : ((c, c), c)$. (Safe inhabitant found by Luke.)
\item
$\sigma(1,3,4)$
is inhabited by the safe term
$$ \lambda x^{(\tau(1,a) \typear \tau(3,b))} y^{(\tau(3,b) \typear \tau(4,c))} z^{\tau(1,c)} w^{\tau(3,c)} . y (x (\lambda u^a . u)) \ .$$

\item $\sigma(2,3,4)$ is inhabited by the safe term
$\lambda x y z w . y (\lambda f . f (\lambda b . b)) (\lambda g . g (\lambda c . c))$
where
$f : ((b, b), b)$,
$g : ((c, c), c)$.

\item It is easy to show that $\sigma(0,2,0)$ (order $4$) is not safely inhabited:
Its only inhabitant up to $\beta$, $\eta$, $\alpha$ conversion
is $\lambda x y z w. y (x z) $ which is unsafe.
Same for $\sigma(0,2,4)$.

\item Luke's example: $(a, (((a, b), b),c), c)$ (order $4$). Its only inhabitant
up to $\beta$, $\eta$, $\alpha$ conversion is $\lambda x.\lambda y. y(\lambda z.zx)$ which is unsafe.

\item At order $3$: The type $( ((a, c), b),  ((c, b), a),  a)$ is inhabited by only one term: $\lambda f g . g (\lambda x . f (\lambda y . c))$ which is unsafe.

\item At order $3$ with only $2$ atoms:
 $( ((b, a), b),  ((a, b), a),  a)$ is only inhabited by the following family of unsafe terms:
 \begin{align*}
& \lambda f g . g (\lambda x_1 . f (\lambda y_1 . x_1)) \\
&\lambda f g . g (\lambda x_1 . f (\lambda y_1 . g (\lambda x_2 . y_1))) \\
&\lambda f g . g (\lambda x_1 . f (\lambda y_1 . g (\lambda x_2 . f (\lambda y_2 . x_i))) \qquad\mbox{where $i = 1, 2$} \\
&\lambda f g . g (\lambda x_1 . f (\lambda y_1 . g (\lambda x_2 . f (\lambda y_2 . g (\lambda x_3 . y_i))) \qquad\mbox{where $i = 1, 2$} \\
&\ldots
\end{align*}
\item At order $2$ all closed $\beta$-normal terms are safe so there is no unsafe type at this order.
\end{enumerate}


\item Luke's conjecture: "there is no uniquely unsafe type that is generated with only two atoms."


\item Conjecture: Even for terms that are not unsafe but that are intrinsically
safe, the variable-capture permitting lemma still holds.
\end{enumerate}

\end{document}
