%Daemon> ini=pdflatex
%Daemon> filter=err+warn
%Daemon> custom_args="-synctex=1"
\documentclass[a4paper,draft]{article}[12pt]
\usepackage{gamesem}
\usepackage{todonotes}
\usepackage{amsmath, amsthm, amssymb}
\usepackage[defblank]{paralist}
\usepackage{shadowbox}
\usepackage{manfnt}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{conjecture}[theorem]{Conjecture}

\newtheorem{lemma}{Lemma}[section]
\newtheorem{proposition}{Proposition}[section]

\theoremstyle{remark}
\newtheorem{remark}{Remark}[section]
\newtheorem{example}{Example}[section]
\newtheorem{property}{Property}[section]
\newtheorem{convention}{Convention}[section]


\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{algorithm}{Algorithm}[section]

% order-decomposition of a 1-stack
\newcommand\orddec{{\sf orddec}}
\newcommand\sig{{\sf sig}}

\newcommand\Nodes{N}% set of nodes
\newcommand\INodes{IN}% set of inner nodes
\newcommand\LNodes{L}% set of leaf nodes

% lambda calculus
\newcommand\typeof[1]{{\sf type}(#1)}
\newcommand\arity{\mathop{\mathrm{arity}}}


\author{William Blum}
\title{Type homogeneity is not a restriction for safe recursion schemes}

\begin{document}
\maketitle
\begin{abstract}

Knapik \emph{et al.} introduced the \emph{safety restriction} for higher-order recursion schemes which constrains both the types and the syntax of the production rules defining a recursion scheme \cite{KNU02}. This constraint gives rise to an equi-expressivity between 
order-$n$ pushdown automata and order-$n$ safe recursion schemes, when such devices are used to generate trees.

We show that the type constraint, called \emph{homogeneity}, is unnecessary in the sense that the syntactic restriction itself is sufficient to obtain the
equi-expressivity result.
\end{abstract}

\section{Introduction}
The concept of \emph{safety} was introduced by Knapik \emph{et al.} in the
context of higher-order recursion schemes and studied further in \cite{KNU02,
demirandathesis}. In the original paper they show that, when used as tree
generators, safe recursion schemes are equivalent to safe pushdown automata
\cite{KNU02}. 
In its original definition, safety consists of two fairly technical constraints:
\begin{enumerate}
\item \textbf{Type-Homogeneity}: This imposes a type constraint on the rules of
the recursion scheme;
\item \textbf{A syntactic restriction} on the relative order of subexpressions
in terms occurring in the right hand side of the recursion scheme rules.
\end{enumerate}

Knapik \emph{et al.}'s result begged a question: what is the automaton
equivalent of `unsafe` recursion schemes?
This question was answered a few years later in a paper introducing a new kind
of automata called Higher-Order Collapsible Pushdown Automata (CPDA), which is
shown to be equivalent to higher-order recursion schemes (whether safe or not)
\cite{hmos-lics08}. More specifically they describe an algorithm that, given an
order-$n$ recursive schemes $G$, construct an equivalent order-$n$ CPDA (in the
sense that they generate the same tree), and conversely. We will refer to the
them as the HMOS conversion procedures.

The safety restriction was further studied in the context of the lambda calculus
in the author's D.Phil thesis \cite{BlumPhd}. Somewhat surprisingly, in the
lambda calculus setting, the type-homogeneity constraint is not necessary to
define a useful calculus. One can define a notion of safe simply-typed lambda
calculus by imposing solely the syntactic constraint above, that we call
\textbf{incremental-binding}, onto the standard simply-typed lambda calculus.
The author's thesis gives expressivity results and present a fully-abstract game
model for this calculus \cite{BlumPhd}. 

This paper reconciles the above observation made in the lambda calculus with the setting of higher-order recursion schemes: we show that if a recursion scheme meets the syntactic criteria of the safety restriction, then regardless of whether the type homogenity is met, the automaton constructed using the HMOS procedure from \cite{hmos-lics08} can be converted into an equivalent order-$n$ (non-collapsible) push-down automaton (PDA). Conversely, given an order-$n$ PDA, the recursion scheme generated by the HMOS transformation induces an equivalent incrementally-bound and homogeneously-typed recursion scheme. 
Consequently, for generating trees, pushdown automata are just as expressive as incrementally-bound recursion schemes.
(In what follows, by abuse of language, we will use ``\emph{incremental-binding}" to refer to the syntactic constraint 2.\ above, even though the concept of binders is foreign to recursion schemes.)

Further, composing the above two transformations yield a methods to 
convert any incrementally-bound recursion scheme into an equivalent \emph{safe} one (in the original sense: incrementally-bound and type-homogeneous). Type-homogeneity is therefore not a proper restriction for safe recursion schemes. This generalizes Knapik \emph{et al.}'s results about  equi-expressivity of pushdown automata and \emph{safe} recursion schemes \cite{KNU02}.
\vspace{1em}

\emph{Note: The result present in this paper was first conjectured in the author's thesis in \cite{BlumPhd}. A partial proof was privately shared by the author in a 2008 note with an incomplete proof for one of the Lemma. Broadbent \cite{Broadbent2009} was the first to observe how to fill the gap by slightly adjusting the definition of stack safety so as to make the inductive proof work. The present paper introduces yet another definition of stack safety which is closer to the original attempt.}

\section{Background}
We first recall some basic notations and recall the definition of higher-order automata and higher-order recursion schemes from the literature \cite{KNU02,hague-collaps-full}.

We consider simple types over a single atom $o$. We call $\Sigma$ a \defname{ranked alphabet} if each symbol $f \in \Sigma$ is assigned a type of the form $o \longrightarrow  \cdots \longrightarrow o \longrightarrow  o$. We write $\arity(f)$ for any typed-term $f$ to denote the arity of its type.

Given a set of typed symbols $S$, the set of \defname{applicative terms} generated from $S$, is defined as the closure of $S$ under the application rule: if $m$ and $n$ are two applicative terms of respective type $A\rightarrow B$ and $A$ then $(m n)$ is also an applicative term of type $B$.

A \defname{$\Sigma$-labelled tree} is a possibly infinite tree where each nodes of the tree is labelled with a symbol in $\Sigma$ with arity matching the number of children nodes in the tree.


\subsection{Higher-order pushdown automata}
\paragraph{Higher-order stacks}
The notion of higher-order stacks is defined inductively. We first fix the base alphabet as a finite set $\Gamma$. An \emph{order-$1$} stack over $\Gamma$ is a sequence of elements of $\gamma \in \Gamma^*$. For $n \in \mathbb{N}$ an \emph{order-$(n+1)$}-stack is defined as a stack of order-$n$ stacks.
The order of a stack $s$ is written $\ord(s)$. The $i^th$ elements in a stack $s$, for $i\geq 0$, is denoted $s_i$ so that a stack $s$ consists of elements in order $s_1$, $\dots$ $s_k$ where $k\geq0$ is called the length of the stack. The stack can then be represented by the notation $[s_1, \cdots s_k]$. The first element $s_1$ and last element $s_k$ are respectively called the `bottom'; and `top' elements of the stack.
The empty stack of order $1$ is denoted $\bot_1$ and represent the stack consisting of no element. We then define the empty $(n+1)$-stack $\bot_{n+1}$ as the singleton sequence $[ \bot_n ]$.

The following operations are defined for an order-$1$ stack $s$:
\begin{eqnarray*}
	push_1^{a} [ \; s_1 \cdots s_m \; ] & := & [ \; s_1 \cdots s_m\, a \; ] \\
	pop_1 [ \; s_1 \cdots s_m s_{m+1} \; ] & : = & [ \; s_1 \cdots s_m \; ] \\
\end{eqnarray*}
and for an order-$(n+1)$ stacks $t$:
\begin{eqnarray*}
	push_{n+1} [ \; t_1 \cdots t_m \; ] & := & [ \; t_1 \cdots t_m\; t_m \; ] \\
	pop_{n+1} [ \; t_1 \cdots t_m t_{m+1} \; ] & : = & [ \; t_1 \cdots t_m \; ] \\
\end{eqnarray*}

The $top_i$ of a stack for $i\geq 1$ is defined as the top $(i-1)$-stack of $s$ (so that $pop_i$ returns $s$ with its top $(i-1)$-stack removed).
Formally, for $s=[ \; s_1 \cdots s_{l+1} \; ]$, $l\geq0$ and $1 \leq i\leq \ord(s)$:
\begin{eqnarray*}
	top_{i} [ \; s_1 \cdots s_{l+1} \; ] & := &  
	\begin{cases}
		s_{l+1} & \mbox{if $i = \ord(s)$} \\
		top_i s_{l+1} & \mbox{if $i < \ord(s)$}
	\end{cases}
\end{eqnarray*}


Following \cite{hmos-lics08}, we define the natural notion of stack prefix: for a given $n$-stack $s$, for any lower-level stack $m$ occurring in $s$, the prefix $s_{\prefixof m}$ of $s$ at $m$ consists of the stack obtained from $s$ by removing all the elements `above' $m$ using a succession of `pop` operations. 

Formally:
\begin{definition}[Stack prefix]
Let $s = [s_1 \cdots s_m]$ be a higher-order stack. Then $s_{\prefixof s_i}$ is defined as $[s_1 \cdots s_i]$ for $1 \leq i \leq m$; and for a stack $t$ occurring in $s_i$ we define $s_{\prefixof t}$ as $[s_1 \cdots {s_i}_{\prefixof t}]$.

Similarly we define the strict stack prefix ${s}{< s_i}$ and $s_{<t}$ respectively as
 $[s_1 \cdots s_{i-1}]$ and $[s_1 \cdots t_{< s_{i}}]$.
\end{definition}

For any operation $\phi$ operating on a higher-order stack and $k\geq1$ we denote by $Phi^k$ the repeated application of $\phi$ $k$ times: $\phi^k = \phi ; \cdots; \phi s$.


\paragraph{Higher-order stacks with links}

We recall the notion of higher-order stack with links where each order-$1$ element is equipped with a link to another stack \cite{hague-collaps-full}:

\begin{definition}[Higher order stack with links (or CPDA stack)]
For a stack alphabet $\Gamma$ and a distinguished bottom-of-stack symbol $\bot\in \Gamma$. An order-$0$ CPDA stack is just a stack symbol. An order-$(n+1)$ CPDA stack $s$ is a non-empty sequence (written $[s_1 \cdots s_l]$) of order $n$ CPDA stacks such that every non-$\bot$ symbol $a$ that occurs in $s$ has a link to a stack of some order $k$ (where $0 \leq k \leq n$) situated below it in $s$; we call the link a $(k + 1)$-link.
\end{definition}

An element of a higher order CPDA stack is written $a^{(j,k)}$ where $a\in \Gamma$ and the exponent $(j,k)$ encodes the pointer associated to the stack symbol. Think of it as a shorthand for the iterated stack operation $pop_j^k$ ($k$ applications of $pop_j$).
The value $j$ is called the order of the link, and $k$ is called the height of the link for $1 \leq j \leq n$ and $k \geq 1$, such that if $j = 1$ then $k = 1$.

\begin{definition}
Let $s$ be a higher-order stack. We define $s^{\langle j \rangle}$ as the operation that replaces
every link occurring in $s$ of the form $(j,k)$ by $(j,k+1)$. Formally,
\begin{align*}
{a^{(j,k)}}^{\langle j \rangle} &= a^{(j,k+1)}   \\
{a^{(j,k)}}^{\langle j' \rangle} &= a^{(j,k)} &   \mbox{when $j\neq j'$,}\\
[s_1 \ldots s_p]^{\langle j \rangle} &= [s_1^{\langle j \rangle} \ldots s_p^{\langle j \rangle}] \ .
\end{align*}
\end{definition}
This operation clearly commutes with prefixing. We will therefore write $s^{\langle j\rangle}_{\prefixof m}$ as an abbreviation for
$(s^{\langle j\rangle})_{\prefixof m} = (s_{\prefixof m})^{\langle j\rangle}$,
for all stack $s$ and stack-symbol $m$ occurring in $s$.


The $top$ and $pop$ operations are defined identically to standard higher-order stacks.
The $push$ operation is defined similarly to the higher-order stacks case except that the push operation now needs to assign a pointer to every element pushed onto the stack. 

In addition to the standard push and pop operations, the CPDA introduces a new $collapse$ operation that `collapses' a given stack to the target of the pointer associated with the top order-$1$ element in the stack. The idea is that if the top-$1$ element is a link from a symbol $a$ to some $(j - 1)$-stack, then the stack can be collapsed to the target of the link by successively performing $k$ times the $pop_j$ operation.

We reproduce here the definition of the $push$ and $collpase$ operations from \cite{hmos-lics08}:

\begin{definition}[Higher-order CPDA operations]
	\label{cpdastackopearations}
We define CPDA operations in terms of the standard stack operations of an order-$n$ PDA with an adequately defined alphabet encoding elements of the form $b^{(o,h)}$ where $b\in\Gamma$ and link indices $o,h \geq 0$. For $1 \leq i  \leq \ord(s)$ and $2  \leq j  \leq ord(s)$:
\begin{align*}
push_1^{b,i}\ s &= push_1^{b^{(i,1)}}\ s \\
collapse\ s &= pop_o^h s  \hbox{\ where } top_1\ s = a^{(o,h)} \\
push_j \underbrace{[ s_1 \ldots s_{l+1} ]}_{s} &=
\left\{
\begin{array}{ll}
\    [s_1\ \ldots\ s_{l+1}\ s_{l+1}^{\langle j \rangle}]  &\hbox{if $j = \ord{s}$;}\\
\    [s_1\ \ldots\ s_{l+1}\ push_j\ s_{l+1}]  &\hbox{if $j<\ord{s}$.}
\end{array}
\right.
\end{align*} 
\end{definition}

\begin{convention}
	In the rest of the paper we will adopt the following abbreviations:
\begin{itemize}
	\item ``$push_1^{a,j}$'' for the operation $push_1\,a^{(j,1)}$;
	\item ``$push_1^a$'' for the operation $push_1\,a^{(j,k)}$ where the components $j$ and $k$ are undetermined.
\end{itemize}
\end{convention}

\begin{convention}[Top stack convention]
\label{conv:top_preserve_links}
In the original definition, the operation $top_i$, that returns the top $(i-1)$-stack of a higher-order stack,
 removes any dangling pointer resulting from the operation. Here, we suppose that $top_i$ is defined in such a way that all pointers are preserved. From an implementation viewpoint, since links are encoded as pairs of integers, this means that $top_i$ just returns an unmodified copy of the top $(i-1)$-stack.
\end{convention}


\begin{definition}
	A tree generating \defname{order-$n$ collapsible pushdown automaton}, $n$-CPDA for short, is a tuple $\langle \Sigma, \Gamma,Q,\delta,q_0 \rangle$ where $\Sigma$ is a ranked alphabet, 
	$\Gamma$ is a stack alphabet, $Q$ is a finite set of states, $q_0$
	is the initial state, and $\delta$ is a transition function $ Q \times \Gamma Γ \longrightarrow Q \times Op + Out$ where
	\begin{itemize}
		\item  $Op$ denotes the set of operations on higher-order stacks with links from Definition \ref{cpdastackopearations}.;
		\item $Out = \{(f, q_1 , \cdots ,q_{\arity(f)})  : f \in \Sigma, q_ i \in Q \}$ denotes the set of possible output emitted at each step of the transition function.
	\end{itemize}
	An order-$n$ CPDA stack is called a \defname{configuration} of an order-$n$ CPDA.
\end{definition}

This definition extends the notion of tree-generating \defname{Pushdown Automaton} (PDA) which are similarly defined on regular higher-order stack with no links and without the $collapse$ stack operation.

\paragraph{Tree generated by a CPDA/PDA}

Think of a CPDA (resp.\ PDA) as a state machine equipped with a higher-order stacks. At each transition, the CPDA can either (i) applies some operation to the stack and update its state, (ii) output the root node $f\in \Sigma$ of the generated tree together with new states for each branch of the tree root. The infinite tree generated by the automaton can then be obtained by recursively spawning new automatons at each child of the node starting with the specified state $q_i$.
The formal definition of the generated tree can be found in \cite{KNU02,hague-collaps-full}.


\paragraph{Higher-order recursion schemes} \hfill


\begin{definition}
	A \defname{higher-order recursion scheme} is a tuple $\langle
	\Sigma, \mathcal{N}, \mathcal{R}, S \rangle$, where $\Sigma$ is a
		ranked alphabet of \emph{terminals};
		$\mathcal{N}$ is a finite set of typed \emph{non-terminals};
		$S$ is a distinguished ground-type symbol of
		$\mathcal{N}$, called the start symbol;
		$\mathcal{R}$ is a finite set of production rules. 
		For each non-terminal $F : (A_1, \ldots, A_n, o) \in \mathcal{N}$ there is exactly one rule of the form:
		$ F z_1 \ldots z_m \rightarrow e$
		where each $z_i$ (called \emph{parameter}) is a
		variable of type $A_i$ and $e$ is an applicative term of type $o$
		generated from the typed symbols in $\Sigma \union \mathcal{N} \union \{z_1:A_1, \ldots, z_m:A_m \}$.

	We say that the recursion scheme is \emph{order-$n$} just in case the order of the highest-order non-terminal is $n$.

\end{definition}

\paragraph{Tree-generated by a recursion scheme}
An applicative term generated from the terminals symbols $\Sigma$ only (without non-terminals), is viewed as a $\Sigma$-labelled tree and called a \defname{value tree}. A recursion scheme defines a (potentially infinite) value tree obtained by unfolding its rewrite rules \emph{ad infinitum}, replacing formal by actual parameters each time, starting from the start symbol $S$. It is formally defined as the least upper bound of the
induced \emph{schematological tree grammar} in the continuous algebra of ranked trees with the appropriate ordering \cite{KNU02,demirandathesis}.

\parpic[r]{
	\raisebox{-30pt}
	{\begin{tikzpicture}[baseline=(root.base),level distance=3ex,inner ysep=0.5mm,sibling distance=13mm]
		\node (root) {$g$}
		child {node {$a$}}
		child {node {$g$}
			child {node {$a$}}
			child {node {$h$}
				child {node {$h$}
					child {node {$\ldots$}}
				}
			}
		};
		\end{tikzpicture}
	}
}
\begin{example}\label{eg:running}
	Let $G$ be the following order-2 recursion scheme:
	\[\begin{array}{rll}
	S & \rightarrow & H \, a\\
	H \, z & \rightarrow & F \, (g \,
	z)\\
	F \, \phi & \rightarrow & \phi \, (\phi \, (F \, h))\\
	\end{array}\]
	with non-terminals $S:o$, $F : ((o, o),o)$, $H:(o,o)$ and terminals $g, h, a$ of arity $2, 1, 0$ respectively.
	Then the tree generated by $G$ is defined by the infinite term
	$g \, a \, (g \, a \, (h \, (h \, (h \,
	\cdots))))$ pictured on the right.%  The only infinite \emph{path} in the
	% tree is the node-sequence $\epsilon \cdot 2 \cdot 22 \cdot 221 \cdot
	% 2211 \cdots$.
	
	%(with the corresponding \textbfit{trace} $g \, g \, h \, h \, h \,
	%\cdots \; \in \; \Sigma^\omega$).
\end{example}


\begin{definition}[Computation tree]
	Comp
\end{definition}
\todo{tree generated by HORS}

\todo{Add \defname{incremental bound computation tree/graph}\cite{blumong:safelambdacalculus}.}

A recursion scheme is \defname{incrementally-bound} just if its computation graph is incrementally-bound.
A recursion scheme is \defname{safe} if it is incrementally-bound and homogeneously-typed.



\subsection{The safety restriction}
\label{sec:safety}

The safety restriction has appeared under different forms in the literature \cite{KNU02,Dam82,demirandathesis,BlumPhd,blumong:safelambdacalculus}.
We include here a reformulation of the definition by Knapik {\it et al.} \cite{KNU02}.

\begin{definition}[Type homogeneity]
\label{subsec:typehomogeneity}
We say that a type $A_1 \longrightarrow \cdots \longrightarrow A_n \longrightarrow o$ with $n \geq 0$ is \defname{homogeneous}
just if $\ord{A_1} \geq \ord{A_2}\geq \cdots \geq \ord{A_n}$, and each $A_1$, \ldots, $A_n$ is homogeneous \cite{KNU02}.
\end{definition}

\begin{definition}[Safe recursion scheme]
Let $G$ be a higher-order recursion scheme whose non-terminals have homogeneous type.
We say that $G$ is \textbf{unsafe} just if it has a production rule $F z_1 \ldots z_m \rightarrow e$ where $e$ contains a subterm that:
\begin{enumerate}
	\item occurs in {\em operand} position in $e$,
	\item contains a parameter of order strictly less than its order.
\end{enumerate}
By ``operand position'' we mean ``in the second position of some
occurrence of the term application operator''.

A recursion scheme is \textbf{safe} if it is not unsafe.
\end{definition}

\begin{remark}
It is possible to define a similar notion of safety for the simply-typed lambda calculus, where the term `safety' gains more sense: performing $\beta$-reduction in this calculus can be safely implemented by capture-permitting substitution without impacting the normal form of a term \cite{BlumPhd,blumong:safelambdacalculus,safety-mirlong2004}.
\end{remark}

We can now state the two main results that are relevant to the result shown in this paper:

\begin{theorem}[Safe HORS - PDA equi-expressivity \cite{KNU02}]
A potentially infinite  $\Sigma$-labelled tree is generated by a \emph{safe} order-$n$ recusion scheme if and only if it is accepted by an order-$n$ pushdown automaton.
\end{theorem}

\begin{theorem}[HORS - CPDA equi-expressivity \cite{hmos-lics08}]
	A potentially infinite $\Sigma$-labelled tree is generated by an order-$n$ recursion scheme if and only if it is accepted by an order-$n$ collapsible pushdown automaton.
\end{theorem}


\section{From recursion scheme to collapsible pushdown automaton}

We fix a higher-order recursion scheme $G$ of order $n$.

\paragraph{Presentation}
We consider the order $n$ collapsible pushdown automaton, obtained from the transformation defined in \cite{hmos-lics08} and denoted $CPDA(G)$, that is constructed from the recursion scheme $G$ \cite[Definition 5.2]{hague-collaps-full}. This automaton is equivalent to $G$ in the sense that they both generate the same tree \cite{hmos-lics08}.

The stack-alphabet $\Gamma$ is given by the set of nodes of the computation graph of $G$ (a tree derived from the long eta-expansion of $G$'s production rules). The automaton proceeds by inductively computing some set of traversals of the computation tree of $G$. One can easily transform $CPDA(G)$ into an automaton that ``prints out'' the traversal that is being computed. This can be done by changing the behaviour of the $push_1$ operation to make it print out the input element before pushing it on the stack. The justification pointers can then be recovered inductively using the node labels: For a variable node, it is the only node-occurrence that binds it in the P-view at that point (which is computable by the induction hypothesis); prime lambda nodes always point to their immediate predecessor; and a non-prime lambda node is always justified by the predecessor of the justifier of the variable node preceding it.

The stack of the current configuration does not suffice in itself to reconstruct the traversal that is being computed
due to the use of some ``destructive'' operations in the CPDA such as $collapse$. Nevertheless, two very useful pieces of information are recoverable from the configuration-stack: the O-view and the P-view of the traversal.

Let $c$ be a configuration. The \defname{long O-view},
\defname{O-view} and \defname{P-view} of the traversal that is currently computed by the configuration $c$, written respectively $\longoview{c}$, $\oview{c}$ and $\pview{c}$, can be recovered as follows:
\begin{compactitem}
\item Long O-view:
\begin{align*}
&  \longoview{s} = \\
 & \quad \left\{
  \begin{array}{ll}
      \epsilon & \mbox{if $top_1 s$ is undefined;} \\
      \longoview{pop_1 s} \cdot top_1 s & \parbox[t]{7cm}{if $top_1\,s \in \INodes_{\sf var}$, $top_1 s$ pointing to its immediate predecesor;} \\
      \longoview{pop_1 s} \cdot top_1 s & \mbox{if $top_1\,s \in \INodes_@$, $@$ having no pointer;} \\
      \longoview{collapse\,s} \cdot top_1 s & \parbox[t]{7cm}{if $top_1\,s \in \INodes_\lambda^{\sf prime}$, $top_1 s$ pointing to its immediate predecesor;} \\
      \longoview{collapse\,s} \cdot top_1 s & \parbox[t]{7cm}{if $top_1\,s \in \INodes_\lambda\setminus \INodes_\lambda^{\sf prime}$, $top_1 s$ pointing to $\ip(\jp(collapse\,s))$.}
    \end{array}
      \right.
\end{align*}
\item The O-view is defined similarly to the long-O-view except that the calculation stops when an @-node is reached:
\begin{align*}
  \oview{s}  &=   top_1 s & \mbox{if $top_1 s \in \INodes_@$, $@$ having no pointer;}
\end{align*}
\item As shown in the original paper, the P-view $\pview{s}$ is given by $top_2\,c$.
\end{compactitem}
\bigskip



\paragraph{Definition}
We now give the formal definition of the collapsible pushdown automaton from the HMOS construction
\cite{hmos-lics08}. The presentation here differs slightly from the original one: In the case (A), when pushing the prime child of an application node $@$ on the stack, we assign it a link pointing to the preceding stack symbol in the top $1$-stack (\ie, the $@$-node itself).
 This modification avoids the case analysis on the value of $j$---the child-index of $u$'s binder---in the
 cases ($V_0$) and ($V_1$), and the sequence of instructions $pop_1^{p+1}$ can just be replaced by
 $pop_1^p ; collapse$. The stack-symbols are the nodes of the computation graph of $G$ and the transition rules are given in
Figure \ref{fig:cpdaprime}. The initial configuration is defined as $c_0 = push_1 \lambda \bot_n$
where $\lambda$ refers to the root (dummy) lambda node of the computation graph of $G$.

The automaton is well-defined in the sense that no collapse can occur at an element whose link is undefined.
(In particular it never occurs at non-lambda nodes.)

\begin{figure}
\begin{center}
\makebox{
\begin{shadowbox}[10cm]
If $u$'s label is not a variables, the action is just a $push_1^v$, where $v$ is an appropriate child of the node $u$. Precisely:
\begin{itemize}
\item $(A)$ If the label is an @ then $\delta(u) = push_1^{E_0(u),{\bf 1}}$.
\item $(S)$ If the label is a $\Sigma$-symbol $f$ then $\delta(u) = push_1^{E_i(u)}$, where $1 \leq i \leq ar(f)$ is the direction requested by the Environment, or Opponent.

\item $(L)$ If the label is a lambda then $\delta(u) = push_1^{E_1(u)}$.
\end{itemize}
Suppose $u$ is a variable which is the $i$-parameter of
its binder and let $p$ be the span of $u$.
\begin{itemize}
\item $(V_1)$ If the variable has order $l\geq 1$, then
$$\delta(u) = push_{n-l+1} ; pop_1^p ; collapse;push_1^{E_i(top_1), n-l+1}$$
\item $(V_0)$ If the variable is of ground type then
$$\delta(u) = pop_1^p ; collapse;push_1^{E_i(top_1)}$$
\end{itemize}
\caption{The transition rules of $CPDA(G)$.}
\label{fig:cpdaprime}
\end{shadowbox}
}
\end{center}
\end{figure}



\paragraph{Reachable configurations}
We use the notion of reachability with respect to the $\rightarrow$-step relation introduced in the original paper:
$c\rightarrow c'$ just if $c' = \delta(top_1 c)(c)$ where $\delta$ is the transition
function defined in \cite[Figure 2]{hague-collaps-full}.
In other words, a configuration is \defname{$\rightarrow$-reachable} if it can be attained starting from the initial configuration $c_0$ by performing
one or more applications of the steps (A), (S), (L), $(V_1)$, $(V_0)$ from the algorithm defining $CPDA(G)$.

The intermediate configurations visited by the internal transitions within a step are therefore not $\rightarrow$-reachable.
A configuration is said to be \defname{reachable} if it is $\rightarrow$-reachable or if it
an intermediate configuration computed during a $\rightarrow$-step
(\ie, if it can be written $(op_1;\ldots;op_k)(c)$ where $c$ is $\rightarrow$-reachable and
$op_1, \ldots, op_k$ are the first $k$ instructions of some $\rightarrow$-step).

\paragraph{Link convention}
Observe that in $CPDA(G)$, when we push a lambda node $\lambda \overline{\xi}$ on the stack, the associated link has order $1$ if it is a prime lambda node (case (A)), and $n-\ord{\lambda \overline{\xi}}+1$ otherwise (case ($V_1$). Hence, since no CPDA instruction can change the link order of an element pushed on the stack, at every stage during the execution of the CPDA the link order can be recovered from the (order of the) node itself.

From now on we will only work with stacks occurring as sub-stack of reachable configurations of $CPDA(G)$
therefore we will omit the order-component altogether when representing stack symbols: we write $\lambda\overline\xi^{k}$ to mean  $\lambda\overline\xi^{(1,k)}$ if $\lambda\overline\xi$ is a prime lambda node
and $\lambda\overline\xi^{(n-\ord{\lambda \overline{\xi}}+1,k)}$ otherwise.

\section{From incrementally-bound RS to PDA}

We now fix an {\bf incrementally-bound} higher-order recursion scheme $G$ of order $n$.
We give a detailed analysis of $CPDA(G)$ that will be used in the next section to derive an equivalent (non-collapsible) order-$n$ pushdown automaton.

\subsection{Incremental order-decomposition}
\paragraph{Observation}
Let $s$ be a 1-stack. For any $l\in \nat$, $s$ can then be written
$$ s = u_{r+1} \cdot \lambda \overline{\eta}_r^{k_r} \cdot u_r \cdot
\ldots \cdot \lambda \overline{\eta}_1^{k_1} \cdot  u_1 $$
where
\begin{itemize}
\item  $\lambda \overline{\eta}_1^{k_1}$ is the
last $\lambda$-node in $s$ with order strictly greater than $l$;

\item for $1 < l \leq r$, $\lambda
\overline{\eta}_l^{k_l}$ is the last $\lambda$-node in $s_{\prefixof
\lambda \overline{\eta}_{l-1}^{k_{l-1}}}$ with order strictly
greater than $\ord{\lambda \overline{\eta}_{l-1}^{k_{l-1}}}$,

\item  $r$ is defined as the smallest number such that
$s_{\prefixof \lambda \overline{\eta}_{r}^{k_{r}}}$ does not contain
any lambda node of order strictly greater than $\lambda
\overline{\eta}_{r}^{k_{r}}$.
\end{itemize}

\noindent In other words:
\begin{itemize}
\item for $1 \leq k \leq r$, all the lambda nodes occurring in $u_l$ have order
strictly smaller than $\ord{\lambda \overline{\eta}_l}$;
\item for $1\leq l<l'\leq r$ we have $\ord{\lambda \overline{\eta}_l^{k_l}}
< \ord{\lambda \overline{\eta}_{l'}^{k_{l'}}}$;
\item $r=0$ if and only if all the lambda node in $s$ have order $\geq l$.
\end{itemize}

The subsequence $\lambda \overline{\eta}_r^{k_r} \ldots \lambda\overline{\eta}_1^{k_1}$ of $s$ consisting of the lambda nodes $\lambda
\overline{\eta}_l^{k_l}$ defined above is called the \defname{incremental order-decomposition of the $1$-stack} $s$ {\bf with respect to} $l\in\nat$.
This sequence is uniquely determined for any given $l\in \nat$.

\smallskip

We now generalize this notion to higher-order stacks.
\begin{definition}
The \defname{incremental order-decomposition of a higher-order stack} $s$ {\bf with respect to $l\in \nat$} (or
order-decomposition for short), written $\orddec_l(s)$, is defined as the order-decomposition of the top order-$1$ stack (defined using convention \ref{conv:top_preserve_links}). Equivalently it can be defined as follows:
\begin{align*}
  \orddec_l(\epsilon) &= \epsilon \\
    \mbox{for } s\neq\epsilon\quad \orddec_l(s) &=     \orddec_{\ord{\lambda\overline{\eta}}}(s_{<\lambda\overline{\eta}}) \cdot \lambda\overline{\eta}^k \\
& \mbox{\quad where } \lambda\overline{\eta}^k \mbox{ is the last node in $top_2\,s$ with order $>l$\ .}
\end{align*}
The \defname{incremental order-decomposition} of $s$, written
$\orddec(s)$, is defined as:
$$
\orddec(s) = \orddec_0(s) \ .
$$
\end{definition}

It follows immediately from the definition that:
\begin{equation}
l<l' \implies \orddec_{l'}(s) \prefixof \orddec_{l}(s)
\label{eqn_safe_prefix}\end{equation}
where $\prefixof$ denotes the sequence-prefix ordering.


\begin{lemma}
\label{lem:push1pop1_orderdecompo} Let $s$ be a (possibly higher-order) stack
such that $top_1\,s \in \INodes_@ \union \INodes_{var}$ and $l\geq 0$.
\begin{enumerate}[(i)]
\item Suppose that $\orddec_l(s) = \langle \lambda
\overline{\eta}_r^{k_r}, \ldots, \lambda \overline{\eta}_1^{k_1}
\rangle$ then for any lambda node $a \in \Gamma$ and link $(j,k)$ we have
 \begin{align*} &\orddec_l(push_1 a^{(j,k)}~s) = \\
&                 \qquad \left\{
                                       \begin{array}{ll}
\orddec_l(s), &\hbox{if $\ord{a}\leq l$;} \\
                                        \langle a^k \rangle, &  \hbox{if $\ord{a} \geq \ord{\lambda \overline{\eta}_r}$}; \\
                                         \langle \lambda \overline{\eta}_r^{k_r}, \ldots, \lambda
\overline{\eta}_i^{k_i}, a^k \rangle,
&                                        \mbox{otherwise,} \\
&                              i = \min \{ i \ \in \{1..r\}| \, \ord{a} <
\ord{\lambda \overline{\eta}_i} \}\ .
\\
                                       \end{array}
                                     \right.
\end{align*}
\item For any non-lambda node $a \in \Gamma$ and link $(j,k)$ we have
$$ \orddec_l(push_1\,a^{(j,k)}\,s) = \orddec_l(s) \ .$$

\item If $top_1 s$ is not a lambda node then
$$ \orddec_l(pop_1\,s) = \orddec_l(s) \ .$$
\end{enumerate}
\end{lemma}
\begin{proof}
Follows immediately from the definition of  $\orddec_l(s)$.
\end{proof}



\begin{lemma}[Incremental binders are in the order-decomposition]
\label{lem:binder_in_ordecompos} Let $c$ be a $\rightarrow$-reachable
configuration of $CPDA(G)$ such that $top_1\,c$ is a variable $x$. Then
\begin{enumerate}[(i)]
\item $\orddec(c)$ contains at least a node with order strictly greater than $\ord(x)$;
\item the last lambda node in $\orddec(c)$ satisfying the first condition is precisely $x$'s binder.
\end{enumerate}
In other words, $x$'s binder is the last lambda node in $\orddec_{\ord x}(c)$.
\end{lemma}
\begin{proof}
\begin{enumerate}[(i)]
\item The top $1$-stack of a $\rightarrow$-reachable configuration contains the P-view of some traversal whose last visited node is the $top_1$ symbol \cite[Corollary 8]{hague-collaps-full}; and
    the P-view of a traversal is exactly the path (in the unfolding of) the
    computation graph from the last visited node to
    the root \cite[Proposition 6]{OngLics2006}. Hence the binder of $x$, whose order
    is strictly greater than $\ord{x}$, occurs in the top $1$-stack.
    Consequently $\orddec(c)$ must contain at least one node of order strictly greater than $l$.

\item Since the recursion scheme is incrementally-bound, $x$'s binder is precisely the first $\lambda$-node in the
 path to the root in the computation tree with order strictly
 greater than $x$. \qedhere
\end{enumerate}
\end{proof}

\subsection{Stack safety}

\begin{definition}[Safe stack]
\label{dfn:safestack} Let $s$ be an order-$j$ non-empty stack for $j\geq1$.

The stack $s$ is \defname{$l$-safe} iff
    \begin{enumerate}[1.]
    \item $\orddec_l(s) = \langle \lambda \overline\eta_r^1, \ldots ,
    \lambda \overline{\eta}_1^1 \rangle$ for some $r\geq 0$ \ie, the height component of the links in $\orddec_l(s)$ are all equal to $1$;

    \item for all $1 \leq q \leq r$ such that $n-\ord{\lambda \overline\eta_q}+1 \leq \ord{s}$:
    \begin{itemize}
    \item for $q = 1$ we have $collapse~s_{\prefixof \lambda\overline\eta_1}$ is $l$-safe;
    \item for $q>1$ we have $collapse~s_{\prefixof \lambda\overline\eta_q}$ is $\ord(\lambda \overline\eta_{q-1})$-safe.
    \end{itemize}
    \end{enumerate}

We say that $s$ is \defname{safe} if it is $0$-safe.
\end{definition}


Since $s$ is a stack, and not necessarily a configuration, it may have dangling pointers.
The condition $n-\ord{\lambda \overline\eta_q}+1 \leq \ord{s}$ in the definition ensures that
$\lambda\overline\eta_j$'s link is not dangling so that we can indeed perform a collapse at $\lambda\overline\eta_j$.

%\begin{lemma}[Equivalent formulation]
%The following formulations of safety are all equivalent:
%\begin{enumerate}
%\item
%Let $s$ be a non-empty stack with order-decomposition
%$\orddec(s)= \langle \lambda \overline\eta_r^{k_r}, \ldots ,
%    \lambda \overline{\eta}_1^{k_1} \rangle$.
%
%Then $s$ is safe iff
%\begin{compactitem}
%    \item if $n-\ord{\lambda \overline\eta_1}+1 \leq \ord{s}$ then
%    $collapse\, s_{\prefixof \lambda\overline\eta_1}$ is $l$-safe;
%    \item  if $r\geq 2$ then  $s_{\prefixof \lambda\overline\eta_2}$ is $\ord{\lambda\overline\eta_1}$-safe.
%\end{compactitem}
%
%\item
%An order-$1$ stack $s$ is safe if and only if $\orddec(s) = \langle \lambda \overline{\eta}_r^1, \ldots , \lambda \overline{\eta}_1^1 \rangle$ with $r\geq 1$;
%a higher-order stack $s$ is safe iff $top_2\,s$ is safe and
%for any lambda node $\lambda\overline\eta \in \orddec(s)$ such that $n-\ord{\lambda \overline{\eta}}+1 \leq \ord{s}$, $collapse\, s_{\prefixof\lambda\overline\eta}$ is $k$-safe where
%$k=l$ if $\lambda\overline\eta$ is the last node in the order decomposition
%otherwise $k$ is the order of the lambda node following
%$\lambda\overline\eta$ in the order decomposition.
%\end{enumerate}
%\end{lemma}

\begin{lemma}
\label{lem:stacksafety_immediate results}
Let $s$ be a stack such that $\orddec_l(s) = \langle \lambda \overline\eta_r^{k_r}, \ldots, \lambda \overline{\eta}_1^{k_1} \rangle$ and $l\geq 0$.
%\begin{enumerate}[(i)]
%\item
If $s$ is $l$-safe and $l \leq k \leq n$ then $s$ is $k$-safe.
%\item $s$ is $l$-safe iif $s_{\prefixof \lambda\overline\eta_1}$ is $l$-safe.
%\end{enumerate}
\end{lemma}
\begin{proof}
Immediate consequence of the definition.
\end{proof}

\begin{lemma}[Collapse simulation]
\label{lem:safecollapsesimulation}
Let $s$ be a sub-stack of a reachable configuration of $CPDA(G)$ and $l\geq 0$.
If $\ord s \geq 2$ and $top_2\,s$ is $l$-safe or if $\ord s = 1$ and $s$ is $l$-safe then for any lambda node $\lambda\overline\eta$ in $\orddec_l(s)$ we have:
$$collapse\,s_{\prefixof\lambda\overline\eta} =
\left\{
\begin{array}{ll}
pop_1\,s_{\prefixof\lambda\overline\eta} & \mbox{if $\lambda\overline\eta$ is prime,}\\
pop_{n-\ord \lambda\overline\eta +1}\,s_{\prefixof\lambda\overline\eta} & \mbox{otherwise.}
\end{array}
\right.
$$
\end{lemma}
\begin{proof}
The $collapse$ operation is defined as  $collapse\,s = pop_j^k\,s$ where $(j,k)\in\nat\times\nat$ is the link attached to $top_1\,s$.
Since $s$ is a sub-stack of a reachable configuration we have $j=1$ if $\lambda\overline\eta$ is prime and $j = n-\ord \lambda\overline\eta +1$ otherwise. Furthermore, since $top_2\,s$ is safe and $\lambda\overline\eta$ belongs to the order-decomposition, the height component necessarily equals $1$.
\end{proof}

\subsubsection{Operations preserving stack safety}
\begin{lemma}
\label{lem:push1pop1_preserves_safety} Let $s$ be a higher-order
stack. Suppose that $s$ is $l$-safe, $l\geq0$. Then:
    \begin{enumerate}[(i)]
    \item $top_{\ord{s}}\,s$ is $l$-safe;
    \item if $top_1\,s$ is not a lambda node then $pop_1\,s$ is $l$-safe;
    \item for every non-lambda node $a$, $1\leq j\leq n$, $k\geq 1$, $push_1\,a^{(j,k)}\,s$ is $l$-safe;
    \item for every lambda node $a$, $push_1\,a^{(1,1)}\,s$ is $l$-safe if $\ord a < l$, and safe if $\ord a \geq l$.
\end{enumerate}
\end{lemma}
\begin{proof}
This is a direct consequence of Lemma \ref{lem:push1pop1_orderdecompo}.
For (vi), the cases $\ord a > l$ and $\ord a < l$ follow immediately from Lemma \ref{lem:push1pop1_orderdecompo}(i);
for $\ord a = l$ it follows from the fact that $\orddec_0 (push_1\,a^{(1,1)}\,s) = \orddec_l (s) \cdot a^1$.
\end{proof}

\begin{lemma}
\label{lem:incrk_qsafe}
Let $0\leq l < n$, $q\geq 0$ and $s$ be a stack of level $1\leq \ord{s} <n$.
If $s$ is $q$-safe then $s^{\langle n-l+1 \rangle}$ is $\max(l,q)$-safe.
\end{lemma}
\begin{proof}
Let $s$ be a safe stack with $1\leq \ord s <n$. We prove the result by induction on the size of $s$.
The base case is the trivial: $s$ is the empty stack.
Step case: Since $s$ is $q$-safe we have
$\orddec_q(s) = \langle \lambda \overline{\eta}_r^1
, \ldots, \lambda \overline{\eta}_1^1   \rangle$.
By (\ref{eqn_safe_prefix}), $\orddec_{\max(l,q)}(s)$ is a prefix of $\orddec_q(s)$.
Let $b$ be the index in $\orddec_q(s)$ of the last node of $\orddec_{\max(l,q)}(s)$:
thus $\lambda\overline\eta_b$ is the last lambda node in $top_2\, s$ with order $>\max(l,q)$.

The stack-operation $(\cdot)^{\langle n-l+1 \rangle}$ updates the pointers as follows:
the height component of each link is incremented if the order of the stack symbol is $l$ and is kept unchanged otherwise.
Hence we have:
\begin{equation*}
\orddec_q(s^{\langle n-l+1 \rangle}) = \langle
\lambda \overline{\eta}_r^1, \ldots,  \lambda \overline{\eta}_{b}^1, \lambda \overline{\eta}_{b-1}^{k}, \lambda \overline{\eta}_{b-2}^1 \ldots,
 \lambda \overline{\eta}_1^{1} \rangle
\end{equation*}
for some $1 \leq k\leq 2$. And therefore:
\begin{equation}
\orddec_{\max(l,q)}(s^{\langle n-l+1 \rangle}) = \langle \lambda \overline{\eta}_r^1
, \ldots, \lambda \overline{\eta}_b^1 \rangle\ . \label{eqn:trivialityofcollapse}
\end{equation}

Now consider an index $j$ such that $b\leq j \leq r$ and $n-\ord{\lambda\overline\eta_j} +1 \leq \ord s$.
Since the height component of $\lambda\overline\eta_j$'s link is not affected by the operation
$(\cdot)^{\langle n-l+1 \rangle}$, this operation commutes with $collapse$ and we have:
\begin{equation}
 collapse\, s^{\langle n-l+1 \rangle}_{\prefixof \lambda \overline{\eta}_{j}}
= (collapse\, s_{\prefixof \lambda \overline{\eta}_{j}})^{\langle n-l+1 \rangle} \ . \label{eqn:commutecollapse}
\end{equation}

By assumption $s$ is $q$-safe therefore $collapse\,s_{\prefixof \lambda \overline{\eta}_{j}}$ is $q$-safe
if $j=b=1$ and $\ord(\lambda\overline\eta_{j-1})$-safe if $j>b\geq1$.

Since $collapse\,s_{\prefixof \lambda\overline\eta_j}$ is strictly smaller than $s$, by the induction hypothesis we have that
$(collapse\,s_{\prefixof \lambda\overline\eta_j})^{\langle n-l+1\rangle}$ is $\max(q,l)$-safe
if $j=b=1$, and $\max(\ord(\lambda\overline\eta_{j-1}),l)$-safe if $j\geq b>1$.

For $j>b$, $\lambda\overline\eta_{j-1}$ occurs in $\orddec_l\, s$ therefore $\ord(\lambda\overline\eta_{j-1}) > l$,
similarly for $j=b$ we have $\ord(\lambda\overline\eta_{j-1}) \leq l$.
Hence $(collapse\,s_{\prefixof \lambda\overline\eta_j})^{\langle n-l+1\rangle}$ is
\begin{enumerate}[(i)]
\item $\max(q,l)$-safe for $j=b=1$,
\item $l$-safe for $j=b>1$, and therefore $\max(q,l)$-safe by Lemma \ref{lem:stacksafety_immediate results},
\item $\lambda\overline\eta_{j-1}$-safe for $j>b$.
\end{enumerate}
This shows that $(collapse\,s_{\prefixof \lambda\overline\eta_j})^{\langle n-l+1\rangle}$ is $\max(l,q)$-safe, and therefore by (\ref{eqn:commutecollapse})
so is $collapse\, s^{\langle n-l+1 \rangle}_{\prefixof \lambda \overline{\eta}_{j}}$.
\end{proof}


\begin{lemma}
\label{lem:cons_qsafety} Let $s$ be a higher-order stack of level $\geq 2$ and $l\geq0$.
 If
\begin{enumerate}[1.]
\item $pop_{\ord{s}}\,s$ is safe,
\item and $top_{\ord{s}}\,s$ is $l$-safe,
\end{enumerate}
then $s$ is $l$-safe.
\end{lemma}
\begin{proof}
Let $s = [s_1 \ldots s_r~s_{r+1}]$ for some $l\geq0$.
We proceed by induction on  $top_{\ord{s}}\,s=s_{r+1}$.
The base case $s_{r+1} = \bot_{\ord{s}-1}$ is trivial.
Suppose that $s_{r+1}$ is not the empty stack.

\begin{enumerate}[(i)]
\item Clearly $\orddec_l\, s = \orddec_l\, s_{r+1}$, hence since $s_{r+1}$ is $l$-safe the lambda nodes in $\orddec_l\, s$ have all a link of height $1$.

\item Let $\lambda \overline{\eta}$ be a lambda node in $\orddec_l(s) = \orddec_l(s_{r+1})$ such that $n-\ord{\lambda \overline{\eta}}+1 \leq \ord{s}$.
Since its link is of height 1 we have
$collapse~s_{\prefixof \lambda \overline{\eta}} = pop_{n-\ord{\lambda \overline{\eta}}+1}~s_{\prefixof \lambda \overline{\eta}}$.

If $n-\ord{\lambda \overline{\eta}}+1 = \ord{s}$ then
$pop_{n-\ord{\lambda \overline{\eta}}+1}~s_{\prefixof \lambda \overline{\eta}} = pop_{\ord{s}}~s_{\prefixof\lambda\overline\eta} = pop_{\ord{s}}~s$ which is $l$-safe by the first assumption
and Lemma \ref{lem:stacksafety_immediate results}.

Otherwise $n-\ord{\lambda \overline{\eta}}+1 < \ord{s}$ and we have:
\begin{align*}
 & collapse\,s_{\prefixof\lambda\overline\eta} \\
&  \qquad    = pop_{n-\ord{\lambda \overline{\eta}}+1}~s_{\prefixof\lambda\overline\eta}
              & \mbox{since $\lambda\overline\eta$'s link has height 1} \\
 & \qquad  = [ s_1 \ldots s_l\ (pop_{n-\ord{\lambda\overline\eta}+1}~ {s_{r+1}}_{\prefixof\lambda\overline\eta}) ]
           & \mbox{$n-\ord{\lambda \overline{\eta}}+1<\ord{s}$}  \\
&  \qquad  = [s_1 \ldots s_p\ (collapse\, {s_{r+1}}_{\prefixof\lambda\overline\eta}) ]
          & \mbox{since $\lambda\overline\eta$'s link has height 1.}
\end{align*}

% we haver to show that  $collapse\,s_{\prefixof \lambda \overline{\eta}}$ is $k$-safe where ...
By the second assumption, $s_{r+1}$ is $l$-safe therefore $collapse~{s_{r+1}}_{\prefixof\lambda\overline\eta}$
is $l$-safe if $\lambda\overline\eta$ is the last node in the $l$-order decomposition,
and $k$-safe where $k$ is the order of the following node in $\orddec_l(s_{r+1})$ otherwise.

Since $|collapse~{s_{r+1}}_{\prefixof\lambda\overline\eta}| < |s_{r+1}|$ we can use the induction hypothesis
to show that the same hold for $[s_1 \ldots s_p\ (collapse\, {s_{r+1}}_{\prefixof\lambda\overline\eta}) ]$. Therefore it is $l$-safe and so is
$collapse\,s_{\prefixof\lambda\overline\eta}$ by the previous equality.
\qedhere
\end{enumerate}
\end{proof}


\begin{lemma}
\label{lem:pushj_safe_implies_l-safe} Let $n>l\geq 1$ and $s$ be a safe higher-order stack such that $2 \leq n-l+1 \leq \ord{s} \leq n$. Then $push_{n-l+1}\ s$ is $l$-safe.
\end{lemma}
\begin{proof}
Let $s=[s_1 \ldots s_{c+1}]$ be a safe higher-order stack such that $2 \leq n-l+1 \leq \ord{s} \leq n$. Then by Lemma \ref{lem:push1pop1_preserves_safety}(i), $s_{c+1}$ is safe.

We show that $push_{n-l+1}~s$ is $l$-safe by finite induction on the order of $s$.
    \begin{compactitem}
      \item Base case: $\ord{s} = n-l+1 $. We have
    $push_{n-l+1}~s = [ s_1 \ldots s_{c+1} s_{c+1}^{\langle n-l+1
    \rangle}]$.

    Since $s_{c+1}$ is safe,  by Lemma \ref{lem:incrk_qsafe} $s_{c+1}^{\langle n-l+1\rangle}$ is $l$-safe, and by Lemma
    \ref{lem:cons_qsafety},  $[ s_1 \ldots s_{c+1} s_{c+1}^{\langle n-l+1
    \rangle}]$ is $l$-safe.

      \item Suppose $\ord{s} > n-l+1$. Then
    $push_{n-l+1}~s = [ s_1 \ldots s_{c+1} push_{n-l+1}\,s_{c+1}]$.

    Since $s_{c+1}$ is safe, by the
    induction hypothesis $push_{n-l+1}~s_{c+1}$ is $l$-safe, and by Lemma \ref{lem:cons_qsafety} so is $[ s_1 \ldots s_{c+1} push_{n-l+1}\,s_{c+1}]$.
\qedhere
    \end{compactitem}
\end{proof}





\subsection{Simulation and proof of correctness}
\begin{proposition}
\label{prop:reachconf_safe}
Let $G$ be an incrementally-bound recursion scheme.
The $\rightarrow$-reachable configurations of $CPDA(G)$ are safe.
\end{proposition}
\begin{proof}
If $n =\ord{c} =1$ then the result holds trivially since $CPDA(G)$ does not contain
any transition of the form $push_j$ for $j>1$ and therefore the links in a reachable configuration all have a height component equal to $1$.

Take $n\geq 2$. We proceed by induction on the $\rightarrow$-step relation. The initial configuration is clearly safe.
Suppose that $c$ is a safe $\rightarrow$-reachable configuration and that
$c \rightarrow c'$. We do a case analysis on $top_1\,c$:
\begin{itemize}
\item (A): We have $c'=push_1^{E_0(u),1}\,c =push_1\,E_0(u)^{(1,1)}\,c$ where $E_0(u)$ denotes a lambda node. It is safe by Lemma \ref{lem:push1pop1_preserves_safety}(iv).

\item (S):
We have $c' = push_1^{a}\,c = push_1\,a^{(j,k)}\,c$ for some dummy lambda node $a$
and undetermined link $(j,k)$. It is safe by Lemma \ref{lem:push1pop1_preserves_safety}(iv) since $\ord a = 0$.

\item (L): We have $c' = push_1^{E_1(u)} = push_1 E_1(u)^{(j,k)}$ where $E_1(u)$ is not a lambda node
 and $j,k\geq1$ are undetermined. It is safe by Lemma \ref{lem:push1pop1_preserves_safety}(iii).

\item ($V_1$) \& ($V_0$): Suppose $u$ is labelled by a variable $x$ of order $l$.
Since $c$ is safe we have $\orddec(c) = \langle \lambda\overline\eta_r^1 , \ldots, \lambda\overline\eta_1^1
\rangle$, $r\geq 0$. Since the recursion-scheme $G$ is safe, by Lemma \ref{lem:binder_in_ordecompos}, $x$'s binder is precisely the last node of $\orddec_l(c)$. Let $b$ be its index in $\orddec(c)$, and $i\geq 1$ be the index of $x$
in $\overline\eta_b$.


\begin{compactitem}
\item  ($V_1$): $l\geq 1$.
We have $c'= push_1\,E_i(top_1)^{(n-l+1,1)}\, t$ where
$t$ is given by $(push_{n-l+1};pop_1^p;collapse)(c) = collapse ((push_{n-l+1}\,c)_{\prefixof \lambda
\overline{\eta}_{b}} )$.
By Lemma \ref{lem:pushj_safe_implies_l-safe} $push_{n-l+1}\, c$ is $l$-safe therefore, since $\lambda\overline\eta_b$ is the last node in $\orddec_l(c)$, by definition of $l$-safety we have that $t$ is $l$-safe. Finally the lambda node $E_i(top_1)$ pushed by the last operation has precisely order $l =\ord(x)$ therefore
\begin{align*}
\orddec_0 (c') &= \langle \lambda\overline\eta_r^1 , \ldots, \lambda\overline\eta_b^1, E_i(top_1(t))^1\rangle \ .
\end{align*}
Thus all the lambda nodes in $\orddec_0 (c')$ have a link of height $1$.

We now need to show that safety is preserved when collapsing at nodes of $\orddec_0 (c')$. Let $b\leq j \leq r$, we have $c'_{\prefixof \lambda\overline\eta_j} =
t_{\prefixof \lambda\overline\eta_j}$. For $j>b$, the l-safety of $t$ implies that $collaspse\, c'_{\prefixof \lambda\overline\eta_j}$ is $\ord{\lambda\overline\eta_{j-1}}$-safe as required. For $j=b$ it gives that $collaspse\, c'_{\prefixof \lambda\overline\eta_b}$ is $l$-safe as required since $l = \ord{E_i(top_1(t))}$.

Now it remains to show that $collapse(c'_{\prefixof{E_i(top_1(t))}}) = collapse\, c'$ is safe.
Since we have $i\geq 1$ the node $top_1(c') = E_i(top_1(t))$ is not a prime lambda node, thus by Lemma \ref{lem:safecollapsesimulation} we can simulate the $collapse$ by a $pop$ of order $n-\ord{E_i(top_1(t))}+1$:
\begin{align*}
collapse\, c' &=  pop_{n-\ord{E_i(top_1(t))}+1}\,c'  \\
               &= pop_{n-l+1}\, c' \\
               &= (push_{n-l+1};pop_1^p;collapse;push_1^{E_i(top_1),n-l+1};pop_{n-l+1})\,c
\end{align*}
The operation $pop_1$ and $push_1$ only affects the top $1$-stack.
Furthermore, since $x$'s binder has order $>l$, its link has order $<n-l+1$ therefore
the collapse operation following the $pop_1^p$ only affects the
top $(n-l)$-stack. Consequently, the operation $pop_{n-l+1}$ effectively restores the configuration
to its value prior to performing the $push_{n-l+1}$ operation:
$$collapse\, c' = c \ .$$
Hence $c'$ is safe.

\item ($V_0$): $l=0$ (which implies that  $b=1$). The configuration $c'$ is given by
$push_1 E_i(top_1)\,collapse(c_{\prefixof \lambda \overline\eta_b})$.
Since $c$ is safe by definition so is $collapse(c_{\prefixof \lambda \overline\eta_b})$.
Since the pushed lambda node $E_i(top_1)$  has order $l=0$, by
Lemma \ref{lem:push1pop1_preserves_safety}(iv) $c'$ is safe.
\qedhere
\end{compactitem}
\end{itemize}
\end{proof}

\begin{definition}[Simulating PDA]
\label{simulating_pda}
Let $G$ be an incrementally-bound recursion scheme.
We define $PDA(G)$ as the higher-order PDA obtained from
$CPDA(G)$ by replacing every $collapse$ operation by $pop_1$ if $top_1\, s$ is prime, and by $pop_{n-\ord{top_1(s)}+1}$ otherwise.
\end{definition}


\begin{theorem}[Correctness of the simulation]
$PDA(G)$ and $CPDA(G)$ are equivalent.
\end{theorem}
\begin{proof}
In $CPDA(G)$, the collapse operation occurs only in the steps ($V_1$) and ($V_0$). For ($V_1$) it is of the form:
$$collapse(pop_1^p(push_{n-l+1}~c))$$
for some $\rightarrow$-reachable configuration $c$, where $top_1\,c$ is a variable $x$ of order $l$ and span $p$.
By the previous proposition, $c$ is safe and by Lemma \ref{lem:pushj_safe_implies_l-safe} $push_{n-l+1}\, c$ is $l$-safe. Since $x$ has span $p$, after the operation $pop_1^p$ the top stack symbol is precisely $x$'s binder, which by Lemma \ref{lem:binder_in_ordecompos}, belongs to $\orddec_l(c)$, therefore
by Lemma \ref{lem:safecollapsesimulation} the collapse can be soundly simulated by a $pop$ of order
$1$ if $x$'s binder is a prime node, and a pop of order $n-\ord(top_1(pop_1^p(push_{n-l+1}\,c))) +1$ otherwise.

The case ($V_0$) is proved similarly.
\end{proof}

\begin{remark}
The same result holds for the original CPDA from the HMOS transformation \cite{hmos-lics08}:
Clearly the two CPDAs have the same set of $\rightarrow$-reachable configurations
so Proposition \ref{prop:reachconf_safe} clearly still holds.
The simulating PDA from \ref{simulating_pda} is obtained from
the CPDA obtained from the HMOS transformation by replacing every $collapse$ operation by $pop_{n-\ord{top_1(s)}+1}$;
and the equality in Lemma \ref{lem:safecollapsesimulation} should be replaced by:
$$collapse\,s_{\prefixof\lambda\overline\eta} = pop_{n-\ord \lambda\overline\eta +1}\,s_{\prefixof\lambda\overline\eta} \ .$$
The correctness of the simulation follows similarly.
\end{remark}

\section{From PDA to incrementally-bound RS}
Let $\mathcal{A}$ be an order-$n$ PDA. Following the HMOS transformation from order-$n$ CPDA to order-$n$ recursion scheme \cite{hmos-lics08} we can produce an equivalent incrementally-bound homogeneously-typed recursion scheme.  Since we do not need to implement the collapse operation we can remove the parameters $\overline\Phi$ altogether from the rules $\mathcal{F}_p^{a,e}$ of the recursion scheme.
The type of the non-terminal $\mathcal{F}_p^{a,e}$  for each stack symbol $a$, $1\leq e \leq n$, and state $1\leq p \leq m$
 becomes:
 $$ \mathcal{F}_p^{a,e} : (n-1)^m \typear \ldots \typear 0^m \typear 0$$
and the production rules are of the general form:
 $$ \mathcal{F}_p^{a,e} \overline{\Psi_{n-1}} \ldots \overline{\Psi_0} \stackrel{(q,\theta)}\rightarrow \Xi_{(q,\theta)}$$
 where the right-hand side is given by:
$$\begin{array}{ll}
\mbox{\bf Cases of $\theta$} & \mbox{\bf Corresponding } \Xi_{(q,\theta)} \\
push_1^{b,k} & \mathcal{F}_q^{b,k} \langle \mathcal{F}_i^{a,e} \overline{\Psi_{n-1}} | i \rangle \overline{\Psi_{n-2}} \ldots
 \overline{\Psi_0} \\
push_j & \mathcal{F}_q^{a,e} \overline{\Psi_{n-1}} \ldots \overline{\Psi_{n-(j-1)}}\langle \mathcal{F}_i^{a,e} \overline{\Psi_{n-1}} \ldots \overline{\Psi_{n-j}} | i \rangle \overline{\Psi_{n-(j+1)}} \ldots
 \overline{\Psi_0} \\
pop_k & \Psi_{n-k,q} \overline{\Psi_{n-k-1}} \ldots \overline{\Psi_{0}}
 \end{array}
 $$

It is easy to verify that this recursion scheme is incrementally-bound and homogeneously-typed.

\bibliographystyle{abbrv}
\bibliography{dphil-all}

\end{document}
