\documentclass{article}

\usepackage{url}
\author{William Blum}
\title{Research student's self-assessment}

\begin{document}


\maketitle
\section*{12 - How has your work gone in the past six months?}

During Hilary term, I attended the course ``Automata Logic \&
Games''.

In March, I attended the Bonn spring school on GAMES and the British
Colloquium in Theoretical Computer Science in Nottingham.

I gave a talk at BCTCS about my MSc dissertation ``Termination
analysis of a subset of CoreML''.

I spent the first four months of my DPhil reading articles in
different directions in order to find an interesting problem to work
on. I have been reading about topics such as program transformation
\cite{compiler-correct-jones-et-al, interpreter-spec-jones}, modal
mu calculus, termination analysis \cite{podelski-transinvar,
podelski-transpredabsrt, abel-structuralrecanalysis}, automata
theory \cite{alur-vpa}.


\section*{13 - Research plan for the next term and vacation:}

I will study a restriction of lambda-calculus called ``safe
lambda-calculus'' as described in \cite{safety-mirlong2004}. My goal
will be to determine some nice properties on safe lambda terms (for
instance whether a safe lambda term is size-change terminating, see
\cite{jones01,blum-mscthesis}).

I will also work on a small project with Matthew Hagues and Luke Ong
about LTL model checking. Using techniques described in
\cite{DBLP:conf/cav/McMillan03, ckos2005, hammer:truly}, we would
like to implement a fast algorithm to verify a property expressed in
LTL on a model given by a Kripke structure.

In July, I will attend the Program Analysis and Transformation
summer school in Copenhagen.

In August, I will attend the international Summer School in
Marktoberdorf.



\section*{14 - Comments}
I am totally unsatisfied with the computer facilities provided by
the department for the research students. The Solaris computers are
obsolete and not at all adapted for office and research needs. The
Ultra 5 machines are extremely slow, almost unusable. The PC
computers are very slow and struggle running Windows Xp.


\bibliography{assessment}
\bibliographystyle{plain}

\end{document}
